\section{Iterative methods for large and sparse linear systems}

Let's consider the following linear system of equations
$$A \mathbf{x} = \mathbf{b}$$
where:

\begin{itemize}
	\item $A \in \mathbb{R}^{n \times n}$
	\item $\mathbf{b} \in \mathbb{R}^{n}$
	\item $\mathbf{x} \in \mathbb{R}^{n}$
	\item $\det(A) \neq 0$
\end{itemize}

n general, direct methods (i.e., methods based on a \textit{manipulation} of $A$) are not suitable whenever $n$ is large, or $A$ is sparse. The average cost of direct methods scales as $n^{3}$. Therefore, we use iterative methods.

We introduce a sequence $\mathbf{x}^{(k)}$ of vectors determined by a recursive relation that identifies the method. In order to initialize the iterative process, it is necessary to provide an initial vector $\mathbf{x}^{(0)}$.

For the method to make sense, it must satisfy the \textbf{convergence property}
$$
	\lim_{ k \to +\infty } \mathbf{x}^{(k)} = \mathbf{x}
$$
Convergence must not depend on the choice of $\mathbf{x}^{(0)}$.

Since the convergence is only guaranteed after an infinite number of iterations,
from a practical point of view we will have to stop the iterative process after a finite number of iterations,
when we believe we have arrived \textit{sufficiently close} to the solution.
For this, we use specific \textbf{stopping criteria}.

Note that, even in exact arithmetic, an iterative method will inevitably be affected by a \textbf{numerical error}.

\subsection{The Jacobi method}

Starting from the $i$-th line of the linear system:
$$
	\sum_{j = 1}^{n} a_{ij} x_{j} = b_{i} \quad \implies \quad a_{i 1} x_{1} + a_{i 2} x_{2} + \dots + a_{i n} x_{n} = b_{i}
$$
Formally, the solution $x_{i}$ for each $i$ is given by
$$
	x_{i} = \frac{b_{i} - \sum_{j \neq i} a_{ij} x_{j}}{a_{ii}}
$$
Obviously we cannot use it because we do not know $x_{j}$ for $j \neq i$. We can introduce an iterative method that updates $x_{i}^{(k+1)}$ using the others $x_{j}^{(k)}$ obtained in the previous step $k$
$$
	x_{i}^{(k+1)} = \frac{b_{i} - \sum_{j\neq i} a_{ij} x_{j}^{(k)}}{a_{ii}} \qquad \forall i = 1, \dots, n
$$
In general, each iteration costs $\sim n^{2}$ operations, so the Jacobi method is competitive if the number of iterations is less that $n$. If $A$ is a sparse matrix, then the cost is only $\sim n$ per iteration.

It should be noted that the solutions $x_{i}^{(k+1)}$ can be computed fully in parallel, which is very competitive for large scale systems.

\subsection{The Gauss-Seidel method}

Let's start from Jacobi's method: $x_{i}^{(k+1)} = \dfrac{b_{i} - \sum_{j \neq i} a_{ij} x_{j}^{(k)}}{a_{ii}}$

At iteration $(k + 1)$, let's consider the computation of $x_{i}^{(k+1)}$. We observe that, for $j < i$ (for $i > 2$), $x_{j}^{(k + 1)}$ is known (we have already calculated it). We can therefore think of using the quantities at step $(k + 1)$ if $j < i$ and (as in the Jacobi method) those at the previous step $k$ if $j > i$.
$$
	x_{i}^{(k+1)} = \frac{b_{i} - \sum_{j<i}a_{i j} x_{j}^{(k+1)} - \sum_{j > i}a_{ij} x_{j}^{(k)}}{a_{ii}}
$$
The computational costs are comparable to those of the Jacobi method. However, unlike the Jacobi, GS is not fully parallelizable.

\subsection{Linear iterative methods}

n general, we consider linear iterative methods of the following form:
$$
	\mathbf{x}^{(k + 1)} = B \mathbf{x}^{(k)} + \mathbf{f} \qquad k \geq 0
$$
where $B \in \mathbb{R}^{n \times n}$, $\mathbf{f} \in \mathbb{R}^{n}$

$B$ is called iteration matrix. Its choice, together with $\mathbf{f}$, uniquely identify the method. So how to choose $B$ and $\mathbf{f}$?

\begin{enumerate}
	\item \textbf{Consistency}: If $\mathbf{x}^{(k)}$ is the exact solution $\mathbf{x}$, then $\mathbf{x}^{(k+1)}$ is again equal to $\mathbf{x}$
	      \begin{equation*}
		      \mathbf{x} = B\mathbf{x} + \mathbf{f} \quad \implies \quad \mathbf{f} = (I - B) \mathbf{x} = (I - B)A^{-1} \mathbf{b}
	      \end{equation*}
	      The former identity gives a relationship between $B$ and $\mathbf{f}$ as a function of the data.
	\item \textbf{Convergence}: Let's introduce the error at step $(k+1)$: $\mathbf{e}^{(k+1)} = \mathbf{x} - \mathbf{x}^{(k+1)}$ and a suitable vector norm $\lVert \cdot \rVert$ (for example, the Euclidian norm).
	      Then we have:
	      \begin{align*}
		      \lVert \mathbf{e}^{(k+1)} \rVert & = \lVert \mathbf{x} - \mathbf{x}^{(k+1)} \rVert = \lVert \mathbf{x} - B \mathbf{x}^{(k)} - \mathbf{f} \rVert   = \qquad \gets \text{consistency} \\[1em]
		                                       & = \lVert \mathbf{x} - B \mathbf{x}^{(k)} - (I - B)\mathbf{x} \rVert = \lVert B \mathbf{e}^{(k)} \rVert                                           \\[1em]
		                                       & \leq \lVert B \rVert \lVert \mathbf{e}^{(k)} \rVert
	      \end{align*}
	      By recursion we obtain
	      \begin{align*}
		      \lVert  \mathbf{e}^{(k-1)} \rVert & \leq \lVert B \rVert \ \lVert B \rVert \ \lVert e^{(k+1)} \rVert                                                                                                   \\[1em]
		                                        & \leq \lVert B \rVert \ \lVert B \rVert \ \lVert B \rVert  \ \lVert \mathbf{e}^{(k-2)} \rVert \leq \dots \leq \lVert B \rVert ^{k+1} \lVert \mathbf{e}^{(0)} \rVert
	      \end{align*}
	      \begin{equation*}
		      \implies \lim_{ k \to \infty } \lVert \mathbf{e}^{(k+1)} \rVert \leq \left(\lim_{ k \to \infty } \lVert B \rVert ^{k+1} \right) \lVert \mathbf{e}^{(0)} \rVert
	      \end{equation*}
	      If $\displaystyle \lVert B \rVert < 1 \implies \lim_{ k \to \infty } \lVert \mathbf{e}^{(k+1)} \rVert = 0$. Therefore, $\lVert B \rVert < 1$ is a \textbf{sufficient condition for convergence}.
\end{enumerate}

\subsubsection{necessary and sufficient condition for convergence}

We define $\rho(B) = \max_{j} \lvert \lambda_{j}(B) \rvert$,
where $\lambda_{j}(B)$ are the eigenvalues of $B$.
$\rho(B)$ is called \textbf{spectral radius} of $B$.
Remember that if $B$ is SPD, then $\rho(B) = \lVert B \rVert$.

\textbf{\textit{Theorem}}

A consistent iterative method with iteration matrix $B$ converges \textbf{if and only if} $\rho(B) < 1$.

\subsubsection{Convergence of Jacobi (J) and Gauss-Seidel (GS) methods}

Let
$$
	A = \begin{bmatrix}
		\ddots &   & -F     \\
		       & D &        \\
		-E     &   & \ddots
	\end{bmatrix}
$$

\begin{itemize}
	\item $D$: diagonal part of $A$
	\item $-E$: lower triangular part of $A$
	\item $-F$: upper triangular part of $A$
\end{itemize}

The Jacobi method can be rewritten as
$$
	D \mathbf{x}^{(k+1)} = (E+F)\mathbf{x}^{(k)} + \mathbf{b}
$$
its iteration matrix is given by
$$
	B_{J} = D^{-1}(E+F) = F^{-1}(D - A) = I - D^{-1}A
$$

For the Gauss-Seidel method we have
$$
	(D - E)\mathbf{x}^{(k+1)} = F \mathbf{x}^{(k)} + \mathbf{b}
$$
The iteration matrix is given by
$$
	B_{GS} = (D - E)^{-1}F
$$

Both methods are consistent.

\begin{itemize}
	\item If $A$ is strictly diagonally dominant by rows/columns, then J and GS converge.
	\item If $A$ is SPD then the GS method is convergent.
	\item If $A$ is tridiagonal, it can be shown that $\rho^{2}(B_{J}) = \rho(B_{GS})$. Therefore both methods converge or fail to converge at the same time. If they converge, GS is faster than J.
\end{itemize}

\subsection{Stopping criteria}

A practical test is needed to determine when to stop the iteration.
The idea is to terminate the iterations when $\dfrac{\lVert \mathbf{x} - \mathbf{x}^{k} \rVert}{\lVert \mathbf{x}^{k} \rVert} \leq \varepsilon$,
where $\varepsilon$ is a user-defined tolerance.
Unfortunately, the error is not known. We can use two criterion:

\begin{enumerate}
	\item \textbf{Residual-based stopping criterion}.
	      $$
		      \frac{\lVert \mathbf{x} - \mathbf{x}^{(k)} \rVert }{\lVert \mathbf{x} \rVert }
		      \leq
		      K(A) \frac{\lVert \mathbf{r}^{(k)} \rVert }{\lVert \mathbf{b} \rVert }
		      \quad \implies \quad
		      \boxed{\frac{\left\lVert  \mathbf{r}^{k} \right\rVert }{\lVert \mathbf{b} \rVert } \leq \varepsilon}
	      $$
	      This is a good stopping criterion whenever $K(A)$ is "small".
	\item \textbf{Distance between consecutive iterates}.
	      Define the relative residual $\boldsymbol{\delta}^{(k)} = \mathbf{x}^{(k+1)} - \mathbf{x}^{(k)}$
	      $$
		      \boxed{\lVert \boldsymbol{\delta}^{(k)} \rVert \leq \varepsilon}
	      $$
	      It can be shown that the relation between the true error and $\boldsymbol{\delta}^{(k)}$ is
	      $$
		      \lVert \mathbf{e}^{(k)} \rVert \leq \frac{1}{1 - \rho(B)} \lVert \boldsymbol{\delta}^{(k)} \rVert
	      $$
	      This is a good stopping criterion only if $\rho(B) \ll 1$.
\end{enumerate}

\subsection{The stationary Richardson method}

Given $\mathbf{x}^{(0)}$, $\alpha \in \mathbb{R}$:
$$
	\mathbf{x}^{(k+1)} = \mathbf{x}^{(k)} + \alpha \underbrace{(\mathbf{b} - A \mathbf{x}^{(k)})}_{\text{residual }\mathbf{r}^{(k)}}
$$
The idea is to update the numerical solution by adding a quantity proportional to the residual (e.g., if the residual is large, the solution at step $k$ should be corrected a lot).

The stationary Richardson method is characterized by
$$
	B_{\alpha} = I - \alpha A \qquad \qquad \mathbf{f} = \alpha \mathbf{b}
$$

\subsubsection{Convergence of the stationary Richardson method}

Let $A$ be symmetric and positive definite matrix (SPD),
then the stationary Richardson method is convergent if and only if
$$
	0 < \alpha < \frac{2}{\lambda_{max}(A)}
$$
with the optimal parameter $\alpha_{opt}$:
$$\alpha_{opt} = \frac{2}{\lambda_{min}(A) + \lambda_{max}(A)}$$
and optimal spectral radius $\rho_{opt}$ (maximum convergence speed):
$$
	\rho_{opt} = \frac{K(A) - 1}{K(A) + 1}
$$

\subsection{Preconditioning techniques}

The optimal value $\rho_{opt} = \frac{K(A) - 1}{K(A) + 1}$ expresses the maximum convergence speed that can be attained with the stationary Richardson method.

Badly conditioned matrices ($K(A) \gg 1$) are characterized by a very low convergence rate.
To improve the speed of convergence, we introduce a SPD matrix $P^{-1}$ (called \textit{preconditioner}). Then, solving $A \mathbf{x} = \mathbf{b}$ is equivalent to the following preconditioned system:
$$
	P^{-\frac{1}{2}} A \underbrace{P^{-\frac{1}{2}} \mathbf{z}}_{\mathbf{x}} =  P^{-\frac{1}{2}} \mathbf{b}
$$
where $\mathbf{x} = P^{-\frac{1}{2}} \mathbf{z}$.

Suppose that $P^{-1}$ has real and positive eigenvalues. We apply the stationary Richardson method to $P^{-1} A$, i.e.,
$$
	\mathbf{x}^{(k+1)} = \mathbf{x}^{(k)} + \alpha P^{-1}\left( \mathbf{b} - A \mathbf{x}^{(k)} \right) = \mathbf{x}^{(k)} + \alpha P^{-1} \mathbf{r}^{(k)}
$$
We obtain the same results as in the non preconditioned case, provided we replace $A$ with $P^{-1}A$:

\begin{itemize}
	\item Convergence: $\qquad0 < \alpha < \dfrac{2}{\lambda_{max}(P^{-1} A)}$
	\item Optimal values: $$\qquad \alpha_{opt} = \dfrac{2}{\lambda_{min}(P^{-1} A) + \lambda_{max}(P^{-1} A)} \qquad \qquad \rho_{opt} = \dfrac{K(P^{-1}A) - 1}{K(P^{-1} A) + 1}$$
\end{itemize}

Therefore, if $K(P^{-1} A) \ll K(A)$, we obtain a higher convergence rate with respect to the unpreconditioned case.

To have $K(P^{-1}A) \ll K(A)$ we need that the linear system in $P$ must be "easily solvable". To this aim, $P$ should have a special structure (e.g., diagonal, triangular, ...).

\subsection{The gradient method}

Let $A \in \mathbb{R}^{n}$ be SPD. In this case, solving the linear system $A \mathbf{x} = \mathbf{b}$ is equivalent to minimizing the quadratic function $\Phi : \mathbb{R}^{n} \to \mathbb{R}$
$$
	\Phi(\mathbf{x}) = \frac{1}{2} \mathbf{x}^{T} A \mathbf{x} - \mathbf{x}^{T}\mathbf{b}
$$
As $A$ is SPD, $\Phi(\mathbf{x})$ defines a paraboloid with global minimum in $\mathbf{x}$. Since $\nabla \Phi(\mathbf{x}) = A\mathbf{x} - \mathbf{b}$ (assuming $A$ is symmetric), we have that the minimum ($\nabla \Phi(\mathbf{x}) = \mathbf{0}$) coincides with the solution of $A \mathbf{x} = \mathbf{b}$. The update rule is
$$
	\mathbf{x}^{(k+1)} = \mathbf{x}^{(k)} - \alpha_{k} \nabla \Phi(\mathbf{x}^{(k)})
$$
where $\alpha_{k}$ is the step size. The optimal parameter $\alpha_{k}$ is chosen by asking that
$$
	\frac{\mathrm d \Phi \left( \mathbf x^{(k+1)} \right)}{\mathrm d \alpha_k} = 0
	\qquad \implies \qquad
	\alpha_{k} = \frac{\left( \mathbf{r}^{(k)} \right)^{T} \mathbf{r}^{(k)}}{\left( \mathbf{r}^{(k)} \right)^{T} A \mathbf{r}^{(k)}}
$$

Note that $\nabla \Phi(\mathbf{x}) = A \mathbf{x} - \mathbf{b} = - \mathbf{r}$. Therefore, the residual $\mathbf r^{(k)}$ at step $k$ is
$$
	\mathbf{r}^{(k)}  = \mathbf{b} - A \mathbf{x}^{(k)} = -\nabla\Phi(\mathbf{x}^{(k)})
$$
Also note that the gradient method can be interpreted as a Richardson method with dynamic parameter $\alpha_{k}$.

\begin{algorithm}
	\caption{Gradient method (steepest descent)}
	\begin{algorithmic}
		\State Given $\mathbf x^{(0)}$, compute $\mathbf r^{(0)} = \mathbf b - A \mathbf x^{(0)}$
		\While{stopping criteria}
		\State $\alpha_k = \dfrac{\left( \mathbf r^{(k)} \right)^T \mathbf r^{(k)}}{\left( \mathbf r^{(k)} \right)^T A\mathbf r^{(k)}}$ \Comment{$\approx n$ flops if $A$ is sparse}
		\State $\mathbf x^{(k+1)} = \mathbf x^{(k)} + \alpha_k \mathbf r^{(k)}$ \Comment{$\approx n$ flops}
		\State $\mathbf r^{(k+1)} = (I - \alpha_k A) \mathbf r^{(k)}$ \Comment{$\approx n$ flops}
		\EndWhile
	\end{algorithmic}
\end{algorithm}

Consider $\mathbf e^{(k)} = \mathbf x^{(k)} - \mathbf x$ the error at iteration $k$. We have that the convergence rate of the gradient method is the same as that of stationary Richardson's method with optimal parameter:
$$
	\lVert \mathbf e^{(k + 1)} \rVert_A \leq \dfrac{K(A) - 1}{K(A) + 1} \lVert \mathbf e^{(k)} \rVert_A
	\qquad \implies \qquad
	\lVert \mathbf e^{(k)} \rVert_A \leq \left( \frac{K(A) - 1}{K(A) + 1} \right)^k \lVert \mathbf e^{(0)} \rVert_A
$$


\subsection{The conjugate gradient method}

We introduce a new updating direction $\mathbf{d}^{(k+1)}$ in such a way that it is $A$-conjugate to all the previous directions $\mathbf{d}^{(j)}, \ j \leq k$ (i.e., orthogonal with respect to the scalar product induced by $A$)
$$
	\left( \mathbf{d}^{(k+1)}, \mathbf{d}^{(j)} \right)_{A} = \left( \mathbf{d}^{(k+1)}, A \mathbf{d}^{(j)} \right) = 0 \qquad \qquad \forall j \leq k \quad \begin{matrix}
		\text{conjugate} \\ \text{directions}
	\end{matrix}
$$

\begin{algorithm}
	\caption{Conjugate Gradient (CG) method}
	\begin{algorithmic}
		\State Given $\mathbf x^{(0)}$, compute $\mathbf r^{(0)} = \mathbf b - A \mathbf x^{(0)}$
		\State Set $\mathbf d^{(0)} = \mathbf r^{(0)}$

		\While{stopping criteria}

		\State $a_k = \dfrac{\left( \mathbf d^{(k)} \right)^T \mathbf r^{(k)}}{\left( \mathbf d^{(k)} \right)^T A\mathbf r^{(k)}}$

		\State $\mathbf x^{(k+1)} = \mathbf x^{(k)} + \alpha_k \mathbf d^{(k)}$ \Comment{Update $\mathbf x^{(k+1)}$ along $\mathbf d^{(k)}$}

		\State $\mathbf r^{(k+1)} = \mathbf r^{(k)} - \alpha_k A \mathbf d^{(k)}$ \Comment{Update $\mathbf r^{(k+1)}$}

		\State $\beta_k = \dfrac{ \left( A \mathbf d^{(k)} \right)^T \mathbf r^{(k+1)} }{ \left( A \mathbf d^{(k)} \right)^T \mathbf d^{(k)} }$

		\State $\mathbf d^{(k+1)} = \mathbf r^{(k+1)} - \beta_k \mathbf d^{(k)}$ \Comment{Update $\mathbf d^{(k+1)}$}
		\EndWhile
	\end{algorithmic}
\end{algorithm}

In exact arithmetic, the CG method converges to the exact solution in at most $n$ iterations. At each iteration $k$, the error $\mathbf{e}^{(k)} = \mathbf{x} - \mathbf{x}^{(k)}$ can be bounded by
$$
	\left\lVert \mathbf{e}^{(k)} \right\rVert _{A} \leq \frac{2c^{k}}{1 + c^{2k}} \left\lVert \mathbf{e}^{(0)} \right\rVert _{A} \qquad \text{with} \qquad c = \frac{\sqrt{ K(A) } - 1}{\sqrt{ K(A) } + 1}
$$
We can also use preconditioning on the CG method.

\subsection{Preconditioned Gradient and Conjugate Gradient}

Both methods converge slowly with poorly conditioned matrices. We can introduce a preconditioner in order to accelerate convergence.

Let $A$ and $P$ be SPD. We consider the following preconditioned system.
$$
	\hat A \hat{\mathbf x} = \hat{\mathbf b}
$$
$$
	\underbrace{P^{- \frac 1 2} A P^{- \frac 1 2}}_{\hat A}
	\underbrace{P^{\frac 1 2} \mathbf x}_{\hat{\mathbf x}}
	=
	\underbrace{P^{- \frac 1 2} \mathbf b}_{\hat{\mathbf b}}
$$
The solution to the original problem can be obtained from the solution $\hat{\mathbf x}$ to the original problem:
$$
	\mathbf x = P^{-\frac 1 2} \hat{\mathbf x}
$$
Let $\mathbf e^{(k)} = \mathbf x^{(k)} - \mathbf x$ be the error at iteration $k$. The error bound of the PCG method is
$$
	\lVert \mathbf e^{(k)} \rVert_A
	\leq
	\dfrac{ 2c^k }{ 1 + c^{2k} } \lVert \mathbf e^{(0)} \rVert
	\qquad \text{with} \qquad
	c = \dfrac{\sqrt{K(P^{-1} A)} - 1}{\sqrt{K(P^{-1} A)}}
$$
If the preconditioner is a "good" preconditioner, then
$$
	\frac{\sqrt{K(P^{-1}A)} - 1}{\sqrt{K(P^{-1} A)} + 1}
	<
	\frac{\sqrt{K(A)} -1}{\sqrt{K(A) + 1}}
$$

\begin{algorithm}[H]
	\caption{Preconditioned Conjugate Gradient (PCG) method}
	\begin{algorithmic}
		\State Given $\mathbf x^{(0)}$, compute $\mathbf r^{(0)} = \mathbf b - A \mathbf x^{(0)}$
		\State Set $\mathbf d^{(0)} = \mathbf r^{(0)}$

		\While{stopping criteria}

		\State $a_k = \dfrac{\left( \mathbf z^{(k)} \right)^T \mathbf r^{(k)}}{\left( A \mathbf d^{(k)} \right)^T \mathbf d^{(k)}}$

		\State $\mathbf x^{(k+1)} = \mathbf x^{(k)} + \alpha_k \mathbf d^{(k)}$

		\State $\mathbf r^{(k+1)} = \mathbf r^{(k)} - \alpha_k A \mathbf d^{(k)}$

		\State $P \mathbf z^{(k + 1)} = \mathbf r^{(k + 1)}$ \Comment{Compute the action of the preconditioner $P$ on $\mathbf r^{(k+1)}$}

		\State $\beta_k = \dfrac{ \left( A \mathbf d^{(k)} \right)^T \mathbf r^{(k+1)} }{ \left( A \mathbf d^{(k)} \right)^T \mathbf d^{(k)} }$

		\State $\mathbf d^{(k+1)} = \mathbf r^{(k+1)} - \beta_k \mathbf d^{(k)}$
		\EndWhile
	\end{algorithmic}
\end{algorithm}

\subsection{Krylov-space methods}

For linear iterative methods (with $P = I$ and $\alpha_{k} = 1 \ \forall k$) as those seen before, we have:
\begin{equation}
	\label{eq:linear-iterative-methods}
	\mathbf{x}^{(k+1)} = \mathbf{x}^{(k)} + \mathbf{r}^{(k)} \qquad k \geq 1
\end{equation}
The following recursive relation for the residuals holds
$$
	\mathbf{r}^{(k+1)} = \mathbf{r}^{(k)} - A \mathbf{r}^{(k)} \qquad k \geq 1
$$
From the above identity, it follows by induction that
$$
	\mathbf{r}^{(k)} = p_{k-1} (A) \mathbf{r}^{(0)} \in \mathrm{span} \ \{\mathbf{r}^{(0)}, \ A \mathbf{r}^{(0)}, \ \dots, \ A^{k-1} \mathbf{r}^{(0)} \}
$$
where $p_{r}(z) = (1 - z)^{r}$ is a polynomial of exact degree $r$.

From $\eqref{eq:linear-iterative-methods}$ we have
$$
	\mathbf{x}^{(k)} = \mathbf{x}^{(0)} + \mathbf{r}^{(0)} + \dots + \mathbf{r}^{(k-1)} = \mathbf{x}^{(0)} + p_{k-1}(A) \mathbf{r}^{(0)}
$$
so $\mathbf{x}^{(k)}$ lies in the affine space $\mathbf{x}^{(0)} + \mathrm{span} \ \{ \mathbf{r}^{(0)}, \ A \mathbf{r}^{(0)}, \ \dots, \ A^{k-1} \mathbf{r}^{(0)} \}$ obtained by shifting the subspace of $\mathbf{r}^{k-1}$.

\begin{tcolorbox}[title=Definition: Krylov subspace]
	Given a nonsingular $A\in \mathbb{R}^{n\times n}$ and $\mathbf{y} \in \mathbb{R}^{n}, \ \mathbf{y} \neq 0$, the $k$-th Kyrlov subspace $\mathscr K_{k}(A, \mathbf{y})$ generated by $A$ from $\mathbf{y}$ is
	$$
		\mathscr K_{k}(A, \mathbf{y}) := \mathrm{span} \ (\mathbf{y}, \ A \mathbf{y}, \ \dots, \ A^{k-1} \mathbf{y})
	$$
\end{tcolorbox}

Clearly, it holds
$$
	\mathscr K_{1}(A, \mathbf{y}) \subseteq \mathcal K_{2} (A, \mathbf{y}) \subseteq \dots
$$

We can choose the $k$-th approximate solution $\mathbf{x}^{(k)} \in \mathbf{x}^{(0)} + \mathscr K_{k}(A, \mathbf{r}^{(0)})$. In particular
$$
	\mathbf{x}^{(k)} = \mathbf{x}^{(0)} + p_{k-1}(A) \mathbf{r}^{(0)}
$$
with a polynomial $p_{k-1}(A)$ of exact degree $k-1$.

When applied to large real-world problems, Kyrlov space solvers often converge very slowly, if at all. In practice, they are therefore nearly always used with preconditioning:
$$
	A \mathbf{x} = \mathbf{b} \qquad \iff \qquad P^{-1} A P^{-1}\mathbf{x} = P^{-1} \mathbf{b} \qquad \iff \qquad \hat{A}\hat{\mathbf{z}} = \hat{\mathbf{b}}, \quad \hat{P} \hat{\mathbf{z}} = \mathbf{x}
$$
Applying a preconditioned Kyrlov space solver means applying the method to $\hat{A} \hat{\mathbf{z}} = \hat{b}$.



\subsection{Solving nonsymmetric linear systems iteratively with Krylov space solvers}

\subsubsection{The biconjugate gradiend (BiCG) method}

While CG (for spd A) has mutually orthogonal residuals $\mathbf{r}^{(k)}$ with
$$
	\mathbf{r}^{(k)}
	=
	p_k(A) \mathbf{r}^{(0)}
	\in
	\mathrm{span} \{ \mathbf r^{(0)}, A \mathbf r^{(0)}, \dots, A^k \mathbf r^{(0)} \}
	=
	\mathscr K_{k+1}(A, \mathbf r^{(0)})
$$
BiCG constructs in the same spaces residuals that are orthogonal to a dual Krylov
space spanned by \textit{shadow residuals}
$$
	\tilde{\mathbf{r}}^{(k)}
	=
	p_k(A) \tilde{\mathbf{r}}^{(0)}
	\in
	\mathrm{span} \{ \tilde{\mathbf r}^{(0)}, A^T \tilde{\mathbf r}^{(0)}, \dots, (A^T)^k \tilde{\mathbf r}^{(0)} \}
	=
	\mathscr K_{k+1}(A^T, \tilde{\mathbf r}^{(0)})
	\equiv
	\tilde{\mathscr{K}}_{k+1}
$$
The initial shadow residual $\tilde{\mathbf{r}}^(0)$ can be chosen freely.
So, BiCG requires two matrix-vector multiplications to extend $\mathscr K_k$ and $\tilde{\mathscr K}_k$:
one multiplication by $A$ and one by $A^T$.

\subsubsection{the BiCGSTAB method}

The biconjugate gradient stabilized method (BiCGSTAB) is a
variant of the biconjugate gradient method (BiCG) and has faster
and smoother convergence than the original BiCG.

It is unnecessary to explicitly keep track of the residuals and search directions of BiCG.
In other words, the BiCG iterations can be performed implicitly.
Unlike the original BiCG method, it doed not require multiplication by $A^T$.

\subsubsection{The GMRES method}

The Generalized Minimum Residual method (GMRES) is a projection method.
The method approximates the solution by the vector in a Krylov subspace
with minimal residual. The arnoldi iteration is used to find this vector.

GMRES approximates the exact solution of $A \mathbf x = \mathbf b$ by the vector
$\mathbf x^{(k)} \in \mathbf x^{(0)} + \mathscr K_k(A, \mathbf r^{(0)})$
that minimizes the Euclidian norm of the residual
$\mathbf r^{(k)} = \mathbf b - A \mathbf x^{(k)}$
